Trauma treatment delay

### 2023-09-14 manuscriptlongversion.Rmd
Introduction

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results

Trauma teams are multidisciplinary and operate in these centers. They consist of individuals from the specialties of emergency medicine, surgery, intensive care, nursing and support staff and are lead by a team leader. Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care. Emergency physicians are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.

>>>>

**Trauma systems**

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@elso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].

Trauma teams are multidisciplinary and operate in these centers. In Sweden, they consist of individuals from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader. Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care. Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel. Once in the trauma room, the team works through ABCDE to treat and stabilise the patient following the protocol present at that specific hospital [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
